---
title: "pipeline benchmark from Tolerance study set"
output:
  html_notebook:
    author: "Julien Tap"
    code_folding: hide
    highlight: tango
    mathjax: null
    theme: cerulean
    toc: yes
    toc_float: yes
date: '`r Sys.Date()`'
---


```{r, message=FALSE, warning=FALSE}

devtools::load_all()
library(ggplot2)
library(reshape2)
library(mixOmics)
library(dplyr)
library(magrittr)
library(DirichletMultinomial)

```


## get started

get biom url from Tolerance data package

```{r}

usearch_lotus = system.file("biom", "OTU_lotus_usearchv7_32bit.biom", package="Tolerance")

usearch_qiime = system.file("biom", "otu_table_non_chimeric_usearch_qiime.biom", package="Tolerance")

usearch2_qiime = system.file("biom", "usearch_otu_table_non_chimeric.biom", package="Tolerance") #sent by Heleen 2017-04-06

vsearch_qiime = system.file("biom", "otu_table_non_chimeric_vsearch_qiime.biom", package="Tolerance")

dada_variant  = system.file("biom", "data_project.biom", package="Tolerance")


UwVsearch_qiime = system.file("biom","otutable_usearchwithvsearchref.biom", package="Tolerance")


VwUsearch_qiime = system.file("biom","otutable_vsearchwithusearchref.biom", package="Tolerance")


#vsearch_unremoval = system.file("biom","vsearch_withoutremoval.biom", package="Tolerance")


deblur_filtered = system.file("biom", "deblur_filtered.seqs.noneukaryota.nonchimeras.biom", package = "Tolerance") # sent by Heleen 2018-08-30


```


## Import and tidy qPCR data



```{r, message=FALSE, warning=FALSE}


b = biomformat::read_biom(usearch_lotus)
m = biomformat::sample_metadata(b)


qPCR = system.file("qPCR","Tolérance_qPCR_Bilan repro_version170804_final.xlsx", package="Tolerance")

qPCR = readxl::read_excel(qPCR)

qPCR =
  qPCR[1:6] %>%
  filter(target == "Bacteria_gr") %>%
  select(visit, Moyenne) %>%
  tidyr::separate(visit,c("Subject","V"),sep = "V") %>%
  mutate(Visit = paste0("V",V)) %>%
  select(Subject,Visit,Moyenne)

all_bacteria = 
  
  m %>%
  tibble::rownames_to_column("Subject_bis") %>%
  select(Subject_bis, Subject, Description) %>%
  tidyr::separate(Description,c("Visit","Product","Dose"),sep = "\\.") %>%
  merge(.,qPCR, by=c("Subject","Visit"))


all_bacteria


```





## compute diversity stat

```{r}


diversity_raw_stat = function(biom_url, filter =TRUE){

b = biomformat::read_biom(biom_url)
o = biomformat::biom_data(b)
#o %>% as.matrix %>% t %>% vegan::rarefy(10000)

#o %>% as.matrix %>% t %>% vegan::rrarefy(10000) %>% vegan::diversity(., index="shannon") %>% rbind(shannon = ., o %>% as.matrix %>% t %>% vegan::rrarefy(10000) %>% vegan::estimateR()) %>% t


if(filter){
  
  
  idx = o %>% as.matrix %>% prop.table(2) %>% BiotypeR::noise.removal(percent=0.005) %>% rownames()
  
  o= o[idx,]
  
  
  
}



r = o %>% 
  as.matrix %>% 
  t %>% 
  vegan::rrarefy(5000) 




res = rbind(
simpson = vegan::diversity(r, index=c("simpson")),
shannon = vegan::diversity(r, index=c("shannon")),
read_nb = apply(o,2,sum),
vegan::estimateR(r)
) %>% t

return(res)

}


diversity_pipeline = 

rbind(
  cbind(diversity_raw_stat(usearch_lotus), pipeline = "usearch_lotus"),
  cbind(diversity_raw_stat(usearch_qiime), pipeline = "usearch_qiime"),
  cbind(diversity_raw_stat(usearch2_qiime), pipeline = "usearch2_qiime"),
  cbind(diversity_raw_stat(vsearch_qiime), pipeline = "vsearch_qiime"),
  cbind(diversity_raw_stat(dada_variant) , pipeline = "dada2"),
  cbind(diversity_raw_stat(UwVsearch_qiime) , pipeline = "UwVsearch_qiime"),
    cbind(diversity_raw_stat(VwUsearch_qiime) , pipeline = "VwUsearch_qiime"),
  #cbind(diversity_raw_stat(vsearch_unremoval) , pipeline = "vsearch_unremoval"),
  cbind(diversity_raw_stat(deblur_filtered) , pipeline = "deblur_filtered")
  
)


diversity_pipeline %>% as.data.frame %>% tibble::rownames_to_column()


```

extract metadata and merge with diversity stat

```{r, message=FALSE, warning=FALSE}


b = biomformat::read_biom(usearch_lotus)

m = biomformat::sample_metadata(b)



stat = m %>% dplyr::select(Description) %>% merge(diversity_pipeline,., by="row.names", all.x = TRUE) %>% melt(id.vars=c("Row.names","Description","pipeline"))

stat$value %<>% as.numeric

stat %<>% cbind(., stringr::str_split(stat$Description, pattern="\\.", simplify = TRUE) )

colnames(stat) = c("Subject","Description","pipeline","index","value","Visit","Product","Dose")


stat


```

# alpha diversity plot

```{r, fig.height=9, fig.width=7}

# 
# stat %>% filter(index=="S.obs") %>% 
#   ggplot() + 
#   geom_violin(aes(x=Visit, y=value), draw_quantiles = 0.5 ) + facet_grid(pipeline~Product+Dose, scales = "free_y")
# 


stat %>% filter(index=="S.obs") %>% 
  group_by(Visit, Product, Dose, pipeline) %>% 
  summarize(mean=median(value), q25 = quantile(value,0.25), q75 = quantile(value,0.75)) %>%  
  ggplot() +  geom_line(aes(y=mean, x=Visit, group=pipeline),  col="red") + 
  geom_line(aes(y=q25, x=Visit, group=pipeline), linetype=2) + 
  geom_line(aes(y=q75, x=Visit, group=pipeline), linetype=2) + 
  facet_grid(pipeline~Product+Dose, scales = "free_y") + ylab("S.obs")



stat %>% filter(index=="simpson") %>% 
  group_by(Visit, Product, Dose, pipeline) %>% 
  summarize(mean=median(value), q25 = quantile(value,0.25), q75 = quantile(value,0.75)) %>%  
  ggplot() +  geom_line(aes(y=mean, x=Visit, group=pipeline),  col="red") + 
  geom_line(aes(y=q25, x=Visit, group=pipeline), linetype=2) + 
  geom_line(aes(y=q75, x=Visit, group=pipeline), linetype=2) + 
  facet_grid(pipeline~Product+Dose, scales = "free_y") + ylab("Simpson")


stat %>% filter(index=="shannon") %>% 
  group_by(Visit, Product, Dose, pipeline) %>% 
  summarize(mean=median(value), q25 = quantile(value,0.25), q75 = quantile(value,0.75)) %>%  
  ggplot() +  geom_line(aes(y=mean, x=Visit, group=pipeline),  col="red") + 
  geom_line(aes(y=q25, x=Visit, group=pipeline), linetype=2) + 
  geom_line(aes(y=q75, x=Visit, group=pipeline), linetype=2) + 
  facet_grid(pipeline~Product+Dose, scales = "free_y") + ylab("Shannon")


stat %>% filter(index=="S.chao1") %>% 
  group_by(Visit, Product, Dose, pipeline) %>% 
  summarize(mean=median(value), q25 = quantile(value,0.25), q75 = quantile(value,0.75)) %>%  
  ggplot() +  geom_line(aes(y=mean, x=Visit, group=pipeline),  col="red") + 
  geom_line(aes(y=q25, x=Visit, group=pipeline), linetype=2) + 
  geom_line(aes(y=q75, x=Visit, group=pipeline), linetype=2) + 
  facet_grid(pipeline~Product+Dose, scales = "free_y") + ylab("S.chao1")







stat %>% filter(index=="read_nb") %>% 
  group_by(Visit, Product, Dose, pipeline) %>% 
  summarize(mean=median(value), q25 = quantile(value,0.25), q75 = quantile(value,0.75)) %>%  
  ggplot() +  geom_line(aes(y=mean, x=Visit, group=pipeline),  col="red") + 
  geom_line(aes(y=q25, x=Visit, group=pipeline), linetype=2) + 
  geom_line(aes(y=q75, x=Visit, group=pipeline), linetype=2) + 
  facet_grid(pipeline~Product+Dose, scales = "free_y") + ylab("read_nb")





```


## association between richness and all bacteria


```{r}

all_bacteria

```



```{r, message=FALSE, warning=FALSE}

stat_qpcr = 
  stat %>%
  merge(.,all_bacteria %>% select(-Subject), by.x=c("Subject","Visit","Product","Dose") , by.y=c("Subject_bis","Visit","Product","Dose"))


stat_qpcr %>%
  #filter(index != "read_nb") %>%
  #filter(index != "se.chao1") %>%
  #filter(index != "se.ACE") %>%
  filter(index == "S.obs") %>%
  ggplot() + 
  geom_point(aes(x=Moyenne,y=value)) + 
  facet_wrap(pipeline~index+Product+Dose, scales = "free") +
  xlim(10,12)



(stat_qpcr %>%
  group_by(pipeline,index,Visit,Product,Dose) %>%
  summarise(rho = cor(value,Moyenne, method="spearman"), n=n(), p.value=cor.test(value,Moyenne, method="spearman") %>% .$p.value) %>%
  arrange(desc(rho)) %>%
  filter(index == "S.obs") -> res)

summary(lm(aov(data=res,  formula = rho~0+Visit+Dose)))




```






## focus on read number


```{r}

stat$value = as.numeric(as.character(stat$value))

stat %>% filter(index=="read_nb", pipeline=="vsearch_qiime") %>% 
  group_by(Visit, Product) %>% 
  summarize(median=median(value), q25 = quantile(value,0.25), q75 = quantile(value,0.75), min=min(value), max=max(value))


```




# alpha diversity correlation



```{r}

stat %>%  dcast(Visit+Product+Dose+Description+index+Subject~pipeline, value.var = "value") %>% filter(index=="S.chao1") %>% ggplot() + geom_point(aes(x=vsearch_qiime, y=usearch_qiime, col=Dose)) + facet_wrap(~Visit+Product)


stat %>% dcast(Visit+Product+Dose+index+Subject~pipeline, value.var = "value") %>% 
  filter(index=="S.obs") %>% 
  group_by(Visit,Product,Dose,index) %>% 
  do(melt(cor(.[,-c(1:5)], method="spearman"))) %>% 
  ggplot() + geom_tile(aes(x=Var1, y=Var2, fill=value)) + 
  facet_grid(Visit~Product+Dose) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0)) +
  xlab("pipeline") + ylab("") + scale_fill_gradient("Alpha_div\nCor.")
  
  
  
  
```


# read mass and genus level plot

Warning: some of the pipeline might use a different taxonomic databases.

```{r, message=FALSE, warning=FALSE}



get_genus_from_biom = function(biom_file){
  
  
  b     = biomformat::read_biom(biom_file)
  otu   = biomformat::biom_data(b)
  tax   = biomformat::observation_metadata(b)
  genus = apply(otu, 2, tapply, paste(tax[,1],tax[,2] ,tax[,3],tax[,4],tax[,5],tax[,6], sep="; "), sum)

  return(genus)  
  
}

genus_dada = get_genus_from_biom(dada_variant)

genus_usearch_qiime = get_genus_from_biom(usearch_qiime)

genus_usearch2_qiime = get_genus_from_biom(usearch2_qiime)

genus_vsearch_qiime = get_genus_from_biom(vsearch_qiime)

genus_usearch_lotus = get_genus_from_biom(usearch_lotus)

genus_deblur_filtered = get_genus_from_biom(deblur_filtered)

#row.names(genus_usearch_lotus) %>% grep("Bifidobacterium|Faecalibacterium|Bacteroides|Lactobacillus|Akkermansia",., value=TRUE)


all_pipeline = 
rbind(
  melt(genus_dada) %>% cbind(pipeline="dada2"),
  melt(genus_usearch_qiime) %>% cbind(pipeline="usearch_qiime"),
  melt(genus_usearch2_qiime) %>% cbind(pipeline="usearch2_qiime"),
  melt(genus_vsearch_qiime) %>% cbind(pipeline="vsearch_qiime"),
  melt(genus_usearch_lotus) %>% cbind(pipeline="usearch_lotus"),
  melt(genus_deblur_filtered) %>% cbind(pipeline="deblur_filtered")
  )


all_pipeline %>% group_by(Var2,pipeline) %>% 
  summarise(read_mass=sum(value)) %>%
  ggplot() + geom_density(aes(x=read_mass, fill=pipeline), alpha=0.8) + scale_x_log10()


all_pipeline %>% group_by(Var2,pipeline) %>% 
  summarise(read_mass=sum(value)) %>% 
  dcast(Var2~pipeline) %>% .[,-1] %>% log10 %>% GGally::ggpairs(progress = F, upper = list(continuous = "cor", corMethod = "spearman")) + xlab("read mass")



all_pipeline %>% group_by(Var2,pipeline) %>% 
  mutate(prop=(value+1)/sum(value+1)) %>% dplyr::select(-value)  %>% 
  filter(grepl("Akkermansia", Var1)) %>%
  dcast(Var2~pipeline,value.var="prop") %>% .[,-1] %>% log10 %>% GGally::ggpairs(progress = F, upper = list(continuous = "cor", corMethod = "spearman")) + xlab("Akkermansia")


all_pipeline %>% group_by(Var2,pipeline) %>% 
  mutate(prop=(value+1)/sum(value+1)) %>% dplyr::select(-value)  %>% 
  filter(grepl("Bifidobacterium", Var1)) %>%
  dcast(Var2~pipeline,value.var="prop") %>% .[,-1] %>% log10 %>% GGally::ggpairs(progress = F, upper = list(continuous = "cor", corMethod = "spearman")) + xlab("Bifidobacterium")




all_pipeline %>% group_by(Var2,pipeline) %>% 
  mutate(prop=(value+1)/sum(value+1)) %>% dplyr::select(-value)  %>% 
  filter(grepl("Methanobrevibacter", Var1)) %>%
  dcast(Var2~pipeline,value.var="prop") %>% .[,-1] %>% log10 %>% GGally::ggpairs(progress = F, upper = list(continuous = "cor", corMethod = "spearman")) + xlab("Methanobrevibacter")



all_pipeline %>% group_by(Var2,pipeline) %>% 
  mutate(prop=(value+1)/sum(value+1)) %>% dplyr::select(-value)  %>% 
  filter(grepl("Bacteroides", Var1)) %>%
  dcast(Var2~pipeline,value.var="prop") %>% .[,-1] %>% log10 %>% GGally::ggpairs(progress = F, upper = list(continuous = "cor", corMethod = "spearman")) + xlab("Bacteroides")


all_pipeline %>% group_by(Var2,pipeline) %>% 
  mutate(prop=(value+1)/sum(value+1)) %>% dplyr::select(-value)  %>% 
  filter(grepl("Prevotellaceae", Var1)) %>% summarise(prop=sum(prop)) %>%
   dcast(Var2~pipeline,value.var="prop") %>% .[,-1] %>% log10 %>% GGally::ggpairs(progress = F, upper = list(continuous = "cor", corMethod = "spearman")) + xlab("Prevotellaceae")




```


# OTUs correlation


```{r}


all_pipeline = 
rbind(
#  melt(genus_dada %>% prop.table(2) %>% BiotypeR::noise.removal(percent=1) ) %>% cbind(pipeline="dada2"),
  #  melt(genus_usearch_lotus %>% prop.table(2) %>% BiotypeR::noise.removal(percent=1)) %>% cbind(pipeline="usearch_lotus"),
  melt(genus_usearch_qiime %>% prop.table(2) %>% BiotypeR::noise.removal(percent=1)) %>% cbind(pipeline="usearch_qiime"),
  melt(genus_usearch2_qiime %>% prop.table(2) %>% BiotypeR::noise.removal(percent=1)) %>% cbind(pipeline="usearch2_qiime"),
  melt(genus_vsearch_qiime %>% prop.table(2) %>% BiotypeR::noise.removal(percent=1)) %>% cbind(pipeline="vsearch_qiime")

  )


all_pipeline %>% #group_by(Var2,pipeline) %>% 
  #mutate(prop=(value+1)/sum(value+1)) %>% dplyr::select(-value) %>%
  ungroup %>%
  mutate(pipeline=paste0(pipeline,"_pp")) %>%
  dcast(Var2~pipeline+Var1,value.var="value") %>%
  dplyr::select(-Var2) %>%
  cor(method="spearman") %>% melt %>% filter(value > 0.9 & value != 1) %>%
  tidyr::separate(Var1,c("pipeline1","tax1"), sep="_pp_", remove = TRUE) %>%
  tidyr::separate(Var2,c("pipeline2","tax2"), sep="_pp_", remove = TRUE) %>%
  dcast(pipeline1~pipeline2) 




all_pipeline_otu = 
rbind(
  #biomformat::read_biom(dada_variant) %>% biomformat::biom_data() %>% as.matrix %>% prop.table(2) %>% BiotypeR::noise.removal(percent=0.1) %>% melt %>% cbind(pipeline="dada2"),
     #biomformat::read_biom(usearch_lotus) %>% biomformat::biom_data() %>% as.matrix %>% prop.table(2) %>% BiotypeR::noise.removal(percent=0.1)%>% melt %>% cbind(pipeline="usearch_lotus"),
   biomformat::read_biom(usearch_qiime) %>% biomformat::biom_data() %>% as.matrix %>% prop.table(2) %>% BiotypeR::noise.removal(percent=0.1) %>% melt %>% cbind(pipeline="usearch_qiime"),
   biomformat::read_biom(usearch2_qiime) %>% biomformat::biom_data() %>% as.matrix %>% prop.table(2) %>% BiotypeR::noise.removal(percent=0.1)%>% melt %>% cbind(pipeline="usearch2_qiime"),
   biomformat::read_biom(vsearch_qiime) %>% biomformat::biom_data() %>% as.matrix %>% prop.table(2) %>% BiotypeR::noise.removal(percent=0.1)%>% melt %>% cbind(pipeline="vsearch_qiime")

  )



all_pipeline_otu %>%
#mutate(prop=(value+1)/sum(value+1)) %>% dplyr::select(-value) %>%
  #ungroup %>%
  mutate(pipeline=paste0(pipeline,"_pp")) %>%
  dcast(Var2~pipeline+Var1,value.var="value") %>%
  dplyr::select(-Var2) %>%
  cor(method="spearman") %>% melt %>% filter(value > 0.9 & value != 1) %>%
  tidyr::separate(Var1,c("pipeline1","tax1"), sep="_pp_", remove = TRUE) %>%
  tidyr::separate(Var2,c("pipeline2","tax2"), sep="_pp_", remove = TRUE) %>%
  dcast(pipeline1~pipeline2)



data.frame(
usearch_qiime = biomformat::read_biom(usearch_qiime) %>% biomformat::biom_data() %>% as.matrix %>% prop.table(2) %>% BiotypeR::noise.removal(percent=0.1) %>% as.matrix %>% dim() %>% .[1],
usearch2_qiime = biomformat::read_biom(usearch2_qiime) %>% biomformat::biom_data() %>% as.matrix %>% prop.table(2) %>% BiotypeR::noise.removal(percent=0.1) %>% as.matrix %>% dim() %>% .[1],
vsearch_qiime = biomformat::read_biom(vsearch_qiime) %>% biomformat::biom_data() %>% as.matrix %>% prop.table(2) %>% BiotypeR::noise.removal(percent=0.1) %>% as.matrix %>% dim() %>% .[1]
)



```

# enterotype

```{r}


et3=DirichletMultinomial::dmn(genus_vsearch_qiime %>% t,k = 3)

et1=DirichletMultinomial::dmn(genus_vsearch_qiime %>% t,k = 1)





```


```{r}

heatmapdmn2(genus_vsearch_qiime %>% t, et1,et3, marges = c(0,1,0,0))

DirichletMultinomial::mixture(et3, assign = TRUE) %>% as.matrix %>% as.data.frame %>% 
  merge(m, by="row.names") %>% dplyr::rename(et = V1) %>% mutate(et = paste0("et",et)) %>%
  tidyr::separate(col=Description, sep="\\.", c("Visit","Group","Dose")) %>%
  ggplot() + geom_bar(aes(x=paste(Group,Dose), fill=et)) + facet_wrap(~Visit)





```








# discriminant analysis with sPLS


```{r, eval=FALSE, include=FALSE}



metadata = m %>% cbind(.,stringr::str_split(m$Description, pattern="\\.", simplify = TRUE)) %>% .[,-c(1,2)] %>% `colnames<-` (c("fastqFile","Subject","Description","Visit","Product","Dose"))

metadata


```





```{r, eval=FALSE, include=FALSE}




res.splsda = 
genus_usearch_qiime %>% add(1) %>%
  prop.table(2) %>% BiotypeR::noise.removal(percent = 0.1) %>% t %>%
  splsda(., 
         Y = data.frame(metadata$Visit,metadata$Product),
         ncomp = 3, multilevel = data.frame(sample=metadata$Subject),
keepX = c(10, 100, 50), logratio="CLR")

design = data.frame(sample=as.factor(metadata$Subject))

genus_usearch_qiime %>% add(1) %>%
  prop.table(2) %>% BiotypeR::noise.removal(percent = 1) %>% t -> df
  
  
  
tune.splsda(X = as.data.frame(df), 
         Y = metadata$Visit, 
         already.tested.X = c(50), 
         ncomp = 2, 
         multilevel = metadata$Subject,
         logratio = "CLR")



plotIndiv(res.splsda, group =metadata$Product , ind.names = metadata$Visit,
legend = TRUE, ellipse = TRUE)


merge(res.splsda$variates$Y, metadata, by="row.names") %>% 
  ggplot() +
  geom_point(aes(x=`comp 1`, y=`comp 2`))



perf(res.splsda)$error.rate

#plotLoadings(res.splsda, comp=1, method = 'median', contrib = "max")

idx = res.splsda$loadings$X[,2] != 0 | res.splsda$loadings$X[,1] != 0 
res.splsda$loadings$X[idx,] %>% as.data.frame

```







```{r}


write.csv2(genus_vsearch_qiime, file="genus_vsearch_qiime_tolerance.csv")


```



